---
title: "Getting Started with nightowl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with nightowl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to FALSE to avoid dependency issues during pkgdown build
)
```

```{r setup}
library(nightowl)
```

# Introduction

**nightowl** is a comprehensive R package for statistical visualization and analysis, specifically designed for clinical research applications. It provides a declarative plotting system built on ggplot2 with R6 classes, survival analysis capabilities, and interactive visualizations.

## Key Features

- **Declarative Plotting System**: R6-based classes for consistent, reproducible visualizations
- **Survival Analysis**: Kaplan-Meier curves and Cox proportional hazards modeling  
- **Interactive Elements**: SVG rendering with hover effects and reactable tables
- **Statistical Summaries**: Built-in statistical tests and summary functions
- **Clinical Focus**: Designed specifically for clinical research workflows

## Installation

```{r}
# Install from GitHub
remotes::install_github("joelgsponer/nightowl")
```

## Basic Usage

### Creating Plots

The core plotting functionality uses the `Plot` R6 class:

```{r}
# Create a basic plot
p <- Plot$new(data = mtcars)
p$add_scatter(x = "wt", y = "mpg")
p$output()
```

### Statistical Summaries

Generate statistical summaries with the `Summary` class:

```{r}
# Create a summary table
summary <- Summary$new(data = mtcars, column = "mpg", method = summarise_numeric_pointrange)
summary$reactable()
```

### Survival Analysis

Perform survival analysis with the `Coxph` class:

```{r}
# Cox proportional hazards model
cox_model <- Coxph$new(data = survival_data)
cox_model$add_variable("treatment")
cox_model$output()
```

## Next Steps

- Explore the [Statistical Summaries](statistical-summaries.html) guide for detailed analysis workflows
- Learn about [Survival Analysis](survival-analysis.html) capabilities
- Browse the [Function Reference](../reference/index.html) for complete API documentation

## Getting Help

- **Documentation**: Complete function reference at https://joelgsponer.github.io/nightowl/
- **Issues**: Report bugs at https://github.com/joelgsponer/nightowl/issues
- **Security**: Review security policy at `inst/SECURITY.md`