---
title: "Getting Started with nightowl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with nightowl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Welcome to nightowl

nightowl is a comprehensive R package designed for statistical visualization and analysis, with a particular focus on clinical research applications. This vignette will get you started with the basic concepts and workflow of nightowl.

## Package Philosophy

nightowl is built around several core principles:

- **Declarative Approach**: Describe what you want, not how to create it
- **Clinical Focus**: Specialized functions for medical and clinical research
- **Performance First**: Built-in caching and optimization for large datasets
- **Interactive by Default**: SVG-based plots with hover effects and interactivity
- **Publication Ready**: Professional styling and export options

## Installation

First, let's install nightowl from GitHub:

```{r installation, eval = FALSE}
# Install from GitHub
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github("joelgsponer/nightowl")
```

Load the package:

```{r setup}
library(nightowl)
library(dplyr)
library(ggplot2)
```

## Your First Plot: Forest Plot

Forest plots are one of nightowl's signature features. Let's start with a simple example:

```{r forest-basic}
# Create a basic forest plot
forestplot(x = 0.75, xmin = 0.5, xmax = 1.1)
```

This creates a forest plot showing an estimate of 0.75 with confidence interval from 0.5 to 1.1. The vertical line at 1.0 represents the null hypothesis line.

### Forest Plot with Custom Styling

```{r forest-styled}
# Create a styled forest plot
forestplot(
  x = 0.65, 
  xmin = 0.4, 
  xmax = 0.9,
  height = 0.4
)
```

## Data Summarization with Summary Class

nightowl provides powerful data summarization capabilities through the `Summary` R6 class. Let's explore this with the built-in `mtcars` dataset:

```{r summary-basic}
# Create a summary of mpg by cylinder count
summary_obj <- Summary$new(
  .data = mtcars, 
  column = "mpg", 
  group_by = "cyl",
  method = summarise_numeric_violin
)

# View the raw summary data
summary_obj$raw()
```

### Visualizing Summaries

The Summary class can automatically create appropriate visualizations:

```{r summary-plot}
# Create summary with point range visualization
mpg_summary <- Summary$new(
  .data = mtcars,
  column = "mpg", 
  group_by = "cyl",
  method = summarise_numeric_violin
)

# Display as interactive table
mpg_summary$reactable()
```

## Declarative Plotting System

nightowl's declarative plotting system lets you describe your visualization at a high level:

```{r declarative-basic}
# Create a declarative plot
# First create a factor column
mtcars$gear_factor <- factor(mtcars$gear)

plot_obj <- plot(
  data = mtcars,
  mapping = list(
    x = "cyl",
    y = "mpg", 
    color = "gear_factor"
  ),
  layers = list(
    list(type = "boxplot"),
    list(type = "generic", geom = "ggplot2::geom_jitter", 
         alpha = 0.6, width = 0.2)
  )
)

plot_obj
```

### Converting Between Formats

One of nightowl's strengths is format flexibility:

```{r format-conversion}
# Convert to ggplot2 object
ggplot_version <- as_ggplot(plot_obj)
class(ggplot_version)

# Convert to HTML (interactive)
# html_version <- as_html(plot_obj)
# This creates an interactive plot with hover effects
```

## Working with Clinical Data

Let's explore nightowl's clinical research capabilities with a simulated dataset:

```{r clinical-data}
# Create simulated clinical trial data
set.seed(123)
clinical_data <- tibble(
  patient_id = 1:200,
  treatment = sample(c("Control", "Treatment"), 200, replace = TRUE),
  age_group = sample(c("Young", "Middle", "Old"), 200, replace = TRUE),
  response = sample(c("No Response", "Response"), 200, replace = TRUE, 
                   prob = c(0.6, 0.4)),
  survival_time = rexp(200, rate = 0.1),
  event = sample(c(0, 1), 200, replace = TRUE, prob = c(0.3, 0.7))
)

head(clinical_data)
```

### Treatment Response Analysis

```{r treatment-response}
# Analyze treatment response by age group
response_summary <- Summary$new(
  .data = clinical_data,
  column = "response",
  group_by = "age_group",
  method = summarise_categorical_barplot
)

response_summary$kable()
```

### Creating Publication-Ready Plots

```{r publication-plot}
# Create a comprehensive treatment comparison plot
treatment_plot <- plot(
  data = clinical_data,
  mapping = list(
    x = "treatment",
    y = "survival_time",
    fill = "treatment"
  ),
  layers = list(
    list(type = "boxplot", alpha = 0.7),
    list(type = "generic", geom = "ggplot2::geom_jitter", 
         alpha = 0.4, width = 0.2)
  )
) +
  theme_minimal() +
  labs(
    title = "Survival Time by Treatment Group",
    subtitle = "Clinical Trial Results",
    x = "Treatment Group", 
    y = "Survival Time (months)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("Control" = "#E74C3C", "Treatment" = "#3498DB"))

treatment_plot
```

## Inline Plots for Reports

nightowl excels at creating small, inline plots perfect for reports and presentations:

```{r inline-plots}
# Create inline histogram
set.seed(42)
sample_data <- rnorm(1000, mean = 100, sd = 15)

inline_hist <- add_inline_plot(
  sample_data,
  style = "Inline-Histogram",
  height = 0.6
)

inline_hist
```

```{r inline-violin}
# Create inline violin plot
inline_violin <- add_inline_plot(
  sample_data,
  style = "Inline-Violin", 
  height = 0.5
)

inline_violin
```

## Performance Features

nightowl includes several performance optimizations:

### Caching System

```{r caching}
# Enable caching for better performance
options(nightowl.cache = TRUE)

# Large dataset example
large_data <- tibble(
  group = sample(LETTERS[1:5], 10000, replace = TRUE),
  value = rnorm(10000),
  category = sample(c("X", "Y", "Z"), 10000, replace = TRUE)
)

# First plot creation (cache miss)
system.time({
  large_plot <- plot(
    data = large_data,
    mapping = list(x = "group", y = "value", color = "category"),
    layers = list(list(type = "boxplot"))
  )
})

# Second access (cache hit - much faster)
system.time({
  large_plot_cached <- plot(
    data = large_data,
    mapping = list(x = "group", y = "value", color = "category"),
    layers = list(list(type = "boxplot"))
  )
})
```

## Interactive Tables

nightowl makes it easy to create professional interactive tables:

```{r interactive-table}
# Create an interactive summary table
summary_table <- clinical_data %>%
  group_by(treatment, age_group) %>%
  summarise(
    n = n(),
    response_rate = mean(response == "Response"),
    median_survival = median(survival_time),
    .groups = "drop"
  )

# Render as interactive table
render_reactable(summary_table)
```

## Styling and Themes

nightowl provides extensive styling options:

```{r styling}
# Load a predefined style
# style_config <- load_style("Boxplot-Dotplot")

# Create styled plot
# First create a factor column
mtcars$cyl_factor <- factor(mtcars$cyl)

styled_plot <- plot(
  data = mtcars,
  mapping = list(x = "cyl_factor", y = "mpg"),
  layers = list(
    list(type = "boxplot"),
    list(type = "dotplot")
  )
)

styled_plot
```

## Next Steps

Now that you've learned the basics, explore these advanced topics:

1. **[Statistical Visualization](statistical-visualization.html)**: Deep dive into nightowl's plotting capabilities
2. **[Advanced Features](advanced-features.html)**: Learn about R6 classes and customization
3. **[Performance Optimization](performance-optimization.html)**: Optimize nightowl for large datasets
4. **[Real-World Examples](real-world-examples.html)**: See nightowl in action with real data

## Key Functions Reference

Here are the main functions you'll use frequently:

| Function | Purpose | Example |
|----------|---------|---------|
| `forestplot()` | Create forest plots | `forestplot(0.5, 0.2, 0.8)` |
| `Summary$new()` | Data summarization | `Summary$new(data, "x", "by")` |
| `plot()` | Declarative plotting | `plot(data, mapping, layers)` |
| `render_reactable()` | Interactive tables | `render_reactable(data)` |
| `add_inline_plot()` | Small inline plots | `add_inline_plot(data, style)` |

## Getting Help

- **Documentation**: Use `?function_name` for detailed help
- **Vignettes**: Browse all vignettes with `browseVignettes("nightowl")`
- **Issues**: Report bugs at [GitHub Issues](https://github.com/joelgsponer/nightowl/issues)
- **Discussions**: Ask questions at [GitHub Discussions](https://github.com/joelgsponer/nightowl/discussions)

Welcome to the nightowl community! 🦉✨