---
title: "Plots as data: the nightowl workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plots as data: the nightowl workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r setup}
library(nightowl)
```

nightowl has one idea at its centre: a plot is a value. A `Plot` wraps a
ggplot together with its rendering options and renders itself to SVG on
demand, once. A `NightowlPlots` vector holds many of them, so a column of a
data frame can be a column of sparklines. Everything else in the package,
the declarative plot specification, the YAML styles, the summary tables,
exists to produce those values and put them in front of a reader.

## Declaring a plot

`DeclarativePlot` builds a ggplot from a specification: data, a mapping of
aesthetics to column names, and an ordered list of layers. Layer types are
the names returned by `nightowl_layers()`.

```{r declare}
p <- DeclarativePlot$new(
  data = ChickWeight,
  mapping = list(x = "Time", y = "weight", colour = "Diet", id = "Chick"),
  layers = list(
    list(type = "traces", alpha = 0.15),
    list(type = "summary", fun.data = "mean_se", geom = "line", linewidth = 1),
    list(type = "summary", fun.data = "mean_se", geom = "pointrange")
  ),
  annotation = list(title = "Chick weight by diet", subtitle = "Individual traces and mean ± SE"),
  svg = list(width = 7, height = 4.5)
)
p$plot
```

The specification is validated before anything is drawn: an unknown field,
mapping key or layer type is an error naming the offending key. Summary
methods and axis zooms are recorded in the plot caption, so a figure
documents how it was made:

```{r captions}
p$captions
```

## Styles

A specification can live in a YAML file. The package ships a few:

```{r styles}
list_styles()
```

`styled_plot()` applies a style to data with the mapping given as
arguments. The style decides the layers; you decide the columns.

```{r styled}
styled_plot(ChickWeight, "Boxplot-SummaryMean", x = "Time", y = "weight", fill = "Diet")$plot
```

```{r raincloud, eval = requireNamespace("ggdist", quietly = TRUE)}
styled_plot(iris, "Raincloud", x = "Species", y = "Sepal.Length", fill = "Species")$plot
```

A style file is a plain description of the layer stack, for example
`Boxplot-SummaryMean.yaml`:

```{r show-style, echo = FALSE, comment = ""}
cat(readLines(system.file("styles", "Boxplot-SummaryMean.yaml", package = "nightowl")), sep = "\n")
```

## Rendering

`$svg()` renders to SVG markup, `$html()` wraps it in a block with the
package stylesheet attached. Rendering is memoised per object. In R Markdown
and Shiny the HTML block can be used directly:

```{r html}
p$html()
```

## Summary tables with inline plots

`Summary` summarises one column per group with a template of calculations.
A calculation can return a number, a one-row data frame, or an inline plot.
The `summarise_*()` templates cover the common cases.

```{r summary}
s <- Summary$new(mtcars, "mpg", group_by = "cyl", method = summarise_numeric_pointrange)
s$raw()
```

The `Pointrange` column is a `NightowlPlots` vector. Table renderers know
what to do with it:

```{r summary-html}
s$html()
```

```{r summary-reactable}
s$reactable()
```

Categorical variables get frequencies and a stacked bar:

```{r categorical}
Summary$new(
  transform(mtcars, gear = factor(gear)), "gear",
  group_by = "cyl", method = summarise_categorical_barplot
)$html()
```

The underlying engine is `calc_summary()`; any named list of functions is a
template:

```{r calc}
calc_summary(
  dplyr::group_by(iris, Species), "Sepal.Length",
  calculations = list(N = length, Mean = format_mean_ci, Hist = add_inline_histogram),
  parameters = list(Hist = list(xlim = range(iris$Sepal.Length)))
)
```

## Colours, theme, options

Three colour-blind-safe palettes ship with the package, and every level
named `"(Missing)"` gets the same grey everywhere.

```{r palettes, fig.height = 2}
pal <- nightowl_palettes()
op <- par(mar = c(0, 4, 0, 0))
plot(NULL, xlim = c(0, 9), ylim = c(0, 3), axes = FALSE, xlab = "", ylab = "")
for (i in seq_along(pal)) {
  rect(seq_along(pal[[i]]) - 1, i - 0.9, seq_along(pal[[i]]), i - 0.1, col = pal[[i]], border = NA)
  mtext(names(pal)[i], side = 2, at = i - 0.5, las = 1)
}
par(op)
```

Defaults are ordinary R options under the `nightowl.` prefix:

```{r options}
nightowl_options()
old <- nightowl_options(palette = "muted")
styled_plot(ChickWeight, "Boxplot", x = "Time", y = "weight", fill = "Diet")$plot
options(old)
```

## Testing figures

Because output is SVG, figures can be compared as text. `expect_snapshot_svg()`
stores a normalised SVG snapshot per operating system and shows changes side
by side in `testthat::snapshot_review()`. For ggplot objects, `vdiffr` works
as usual on `$plot`.
