---
title: "Statistical Summaries and Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Summaries and Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to FALSE to avoid dependency issues during pkgdown build
)
```

```{r setup}
library(nightowl)
library(dplyr)
```

# Statistical Summaries

The nightowl package provides a comprehensive suite of statistical summary functions designed for clinical research applications. These functions follow a consistent API pattern and integrate seamlessly with the plotting system.

## Summary Function Family

The `summarise_*` family of functions provides modular statistical summaries:

### Numeric Summaries

```{r}
# Point range summaries (mean ± CI)
summarise_numeric_pointrange(mtcars, "mpg", group_by = "cyl")

# Forest plot summaries
summarise_numeric_forestplot(mtcars, "mpg", group_by = "am")

# Histogram summaries
summarise_numeric_histogram(mtcars, "mpg", group_by = "vs")

# Violin plot summaries  
summarise_numeric_violin(mtcars, "mpg", group_by = "gear")
```

### Categorical Summaries

```{r}
# Frequency tables
summarise_categorical(mtcars, "cyl", group_by = "am")

# Bar plot summaries
summarise_categorical_barplot(mtcars, "cyl", group_by = "vs")
```

## Summary Class

The `Summary` R6 class provides an object-oriented interface to statistical summaries:

```{r}
# Create a summary object
summary <- Summary$new(
  data = mtcars, 
  column = "mpg", 
  method = summarise_numeric_pointrange
)

# Generate interactive table
summary$reactable()

# Get raw summary data
summary$data
```

## Statistical Tests

Built-in statistical testing with automatic method selection:

```{r}
# Chi-square analysis for grouped data
grouped_chisq(data = mtcars, x = "cyl", y = "am", group_by = "vs")

# Automatic test selection
calc_test(mtcars, "mpg", group_by = "cyl")  # ANOVA for numeric
calc_test(mtcars, "cyl", group_by = "am")   # Chi-square for categorical
```

## Advanced Workflows

### Arranging Multiple Summaries

```{r}
# Create multiple summaries
summaries <- list(
  summarise_numeric_pointrange(mtcars, "mpg", group_by = "cyl"),
  summarise_categorical(mtcars, "am", group_by = "cyl"),
  summarise_numeric_violin(mtcars, "hp", group_by = "cyl")
)

# Arrange in a structured layout
arrange_summaries(summaries, ncol = 2)
```

### Custom Summary Methods

```{r}
# Define custom summary function
custom_summary <- function(data, column, ...) {
  data %>%
    group_by(...) %>%
    summarise(
      n = n(),
      median = median(.data[[column]], na.rm = TRUE),
      iqr = IQR(.data[[column]], na.rm = TRUE),
      .groups = "drop"
    )
}

# Use with Summary class
Summary$new(data = mtcars, column = "mpg", method = custom_summary)
```

## Integration with Plotting

Statistical summaries integrate seamlessly with the plotting system:

```{r}
# Create plot with statistical overlay
p <- Plot$new(data = mtcars)
p$add_violin(x = "factor(cyl)", y = "mpg")
p$add_summary(method = summarise_numeric_pointrange, group_by = "cyl")
p$output()
```

## Clinical Research Applications

### Treatment Effect Analysis

```{r}
# Compare treatment groups
treatment_summary <- summarise_numeric_forestplot(
  data = clinical_data,
  column = "response", 
  group_by = "treatment",
  reference = "control"
)

# Generate forest plot
Plot$new(treatment_summary)$add_forestplot()$output()
```

### Biomarker Analysis

```{r}
# Analyze biomarker distributions
biomarker_summary <- summarise_numeric_violin(
  data = biomarker_data,
  column = "biomarker_level",
  group_by = c("treatment", "timepoint")
)

# Create grouped visualization
Summary$new(biomarker_summary)$reactable()
```

## Best Practices

1. **Consistent Grouping**: Use consistent grouping variables across analyses
2. **Method Selection**: Choose appropriate summary methods for data types
3. **Statistical Testing**: Always include appropriate statistical tests
4. **Visualization**: Combine summaries with plots for comprehensive analysis
5. **Documentation**: Document custom summary methods for reproducibility

## See Also

- [Getting Started](getting-started.html) for basic usage
- [Survival Analysis](survival-analysis.html) for time-to-event analysis
- [Function Reference](../reference/index.html) for complete API documentation