---
title: "Statistical Visualization with nightowl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Visualization with nightowl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  dpi = 300
)
```

## Introduction

nightowl provides a comprehensive suite of statistical visualization tools designed specifically for clinical research and data analysis. This vignette demonstrates the full range of plotting capabilities, from basic visualizations to complex multi-panel displays.

```{r setup}
library(nightowl)
library(dplyr)
library(ggplot2)
library(survival)
```

## Forest Plots: The Heart of Clinical Research

Forest plots are nightowl's flagship feature, essential for meta-analyses and clinical trial reporting.

### Basic Forest Plots

```{r forest-basic}
# Simple forest plot with confidence interval
forestplot(estimate = 0.65, lower = 0.45, upper = 0.95)
```

### Multi-Study Forest Plots

```{r forest-multi}
# Create dataset for multiple studies
studies <- tibble(
  study = paste("Study", LETTERS[1:8]),
  n = sample(50:200, 8),
  estimate = rnorm(8, mean = 0.7, sd = 0.15),
  se = runif(8, 0.08, 0.20)
) %>%
  mutate(
    lower = estimate - 1.96 * se,
    upper = estimate + 1.96 * se
  )

# Create comprehensive forest plot
studies %>%
  forestplot(
    estimate = estimate,
    lower = lower, 
    upper = upper,
    study_labels = study,
    height = 0.6
  ) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", alpha = 0.7) +
  labs(
    title = "Treatment Effect Across Multiple Studies",
    subtitle = "Odds Ratios with 95% Confidence Intervals",
    x = "Odds Ratio"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12)
  )
```

### Subgroup Forest Plots

```{r forest-subgroup}
# Forest plot with subgroup analysis
subgroup_data <- tibble(
  subgroup = rep(c("Age < 65", "Age ≥ 65", "Male", "Female", "Stage I-II", "Stage III-IV"), each = 3),
  category = rep(c("Overall", "Treatment A", "Treatment B"), 6),
  estimate = c(0.75, 0.68, 0.82, 0.71, 0.65, 0.78, 0.73, 0.69, 0.77, 0.69, 0.61, 0.76, 0.72, 0.67, 0.78, 0.74, 0.70, 0.79),
  lower = estimate - 0.15,
  upper = estimate + 0.15
)

subgroup_data %>%
  forestplot(
    estimate = estimate,
    lower = lower,
    upper = upper,
    study_labels = paste(subgroup, "-", category),
    color_by = subgroup
  ) +
  facet_wrap(~subgroup, scales = "free_y", ncol = 2) +
  labs(title = "Subgroup Analysis of Treatment Effects") +
  theme_minimal()
```

## Survival Analysis Visualizations

nightowl excels at survival analysis visualization with specialized functions for Kaplan-Meier curves and Cox regression.

### Kaplan-Meier Survival Curves

```{r km-basic}
# Create simulated survival data
set.seed(42)
surv_data <- tibble(
  patient_id = 1:300,
  treatment = sample(c("Control", "Treatment A", "Treatment B"), 300, replace = TRUE),
  age_group = sample(c("Young", "Middle", "Elderly"), 300, replace = TRUE),
  time = rexp(300, rate = 0.1),
  event = rbinom(300, 1, 0.7)
)

# Basic Kaplan-Meier plot
km_fit <- fit_km(surv_data, time = "time", event = "event", treatment = "treatment")
plot_km(km_fit) +
  labs(title = "Kaplan-Meier Survival Curves by Treatment")
```

### Advanced Survival Plots with Risk Tables

```{r km-advanced}
# Kaplan-Meier with risk tables and p-values
plot_grouped_km(
  surv_data,
  time = "time",
  event = "event", 
  treatment = "treatment",
  show_risk_table = TRUE,
  show_pvalue = TRUE
) +
  theme_minimal() +
  labs(
    title = "Overall Survival by Treatment Group",
    subtitle = "With risk tables and statistical comparison"
  )
```

### Cox Proportional Hazards Visualization

```{r cox-analysis}
# Cox regression with forest plot
cox_model <- fit_coxph(
  surv_data,
  time = "time",
  event = "event",
  covariates = c("treatment", "age_group")
)

plot_coxph(cox_model) +
  labs(
    title = "Cox Proportional Hazards Model",
    subtitle = "Hazard Ratios with 95% Confidence Intervals"
  )
```

## Categorical Data Visualization

### Chi-Square Analysis with Visualization

```{r chisq-analysis}
# Create categorical dataset
categorical_data <- tibble(
  treatment = sample(c("Placebo", "Drug A", "Drug B"), 400, replace = TRUE),
  response = sample(c("No Response", "Partial Response", "Complete Response"), 
                   400, replace = TRUE, prob = c(0.5, 0.3, 0.2)),
  severity = sample(c("Mild", "Moderate", "Severe"), 400, replace = TRUE)
)

# Grouped chi-square analysis
chisq_result <- grouped_chisq(
  categorical_data,
  x = "response",
  by = "treatment"
)

plot_grouped_chisq(chisq_result) +
  labs(
    title = "Treatment Response Analysis",
    subtitle = "Chi-square test with residuals visualization"
  )
```

### Donut Charts for Proportions

```{r donut-plots}
# Create donut plot for response rates
response_summary <- categorical_data %>%
  count(response) %>%
  mutate(percentage = n / sum(n) * 100)

donut_plot(
  response_summary,
  values = "n",
  labels = "response",
  title = "Overall Response Distribution"
) +
  scale_fill_manual(values = c("#E74C3C", "#F39C12", "#27AE60"))
```

## Continuous Data Visualization

### Advanced Boxplot Configurations

```{r boxplot-advanced}
# Enhanced boxplot with multiple comparisons
mtcars %>%
  plot(
    mapping = list(
      x = "factor(cyl)",
      y = "mpg",
      fill = "factor(cyl)"
    ),
    layers = list(
      list(type = "boxplot", alpha = 0.7, outlier.alpha = 0.6),
      list(type = "generic", geom = "ggplot2::geom_jitter", 
           width = 0.2, alpha = 0.4, size = 2)
    )
  ) +
  stat_compare_means(method = "anova", label.y = 35) +
  stat_compare_means(
    comparisons = list(c("4", "6"), c("6", "8"), c("4", "8")),
    method = "t.test"
  ) +
  labs(
    title = "Fuel Efficiency by Cylinder Count",
    subtitle = "Boxplots with statistical comparisons",
    x = "Number of Cylinders",
    y = "Miles per Gallon"
  ) +
  theme_minimal()
```

### Violin Plots with Summary Statistics

```{r violin-plots}
# Violin plots with embedded boxplots
iris %>%
  plot(
    mapping = list(
      x = "Species",
      y = "Sepal.Length",
      fill = "Species"
    ),
    layers = list(
      list(type = "violin", alpha = 0.7, trim = FALSE),
      list(type = "boxplot", width = 0.1, alpha = 0.8),
      list(type = "generic", geom = "ggplot2::stat_summary",
           fun = "mean", geom = "point", size = 3, color = "red")
    )
  ) +
  labs(
    title = "Sepal Length Distribution by Species",
    subtitle = "Violin plots with embedded boxplots and mean indicators"
  ) +
  theme_minimal()
```

## Multi-Panel and Complex Visualizations

### Correlation Matrices

```{r correlation-matrix}
# Create correlation matrix visualization
library(GGally)

mtcars %>%
  select(mpg, disp, hp, drat, wt, qsec) %>%
  ggpairs(
    title = "Motor Car Characteristics Correlation Matrix",
    theme = ggpairs_theme()
  )
```

### Faceted Analysis

```{r faceted-analysis}
# Multi-panel analysis by groups
iris %>%
  plot(
    mapping = list(
      x = "Sepal.Width",
      y = "Sepal.Length",
      color = "Species"
    ),
    layers = list(
      list(type = "scatter", size = 2, alpha = 0.7),
      list(type = "smooth", method = "lm", se = TRUE)
    )
  ) +
  facet_wrap(~Species, scales = "free") +
  labs(
    title = "Sepal Dimensions Relationship by Species",
    subtitle = "Scatter plots with linear regression fits"
  ) +
  theme_minimal()
```

## Interactive Visualizations

### Interactive Tables with Statistical Summaries

```{r interactive-tables}
# Comprehensive summary table
summary_table <- categorical_data %>%
  group_by(treatment, severity) %>%
  summarise(
    n = n(),
    complete_response_rate = mean(response == "Complete Response"),
    partial_response_rate = mean(response == "Partial Response"),
    overall_response_rate = mean(response %in% c("Complete Response", "Partial Response")),
    .groups = "drop"
  ) %>%
  mutate(
    across(ends_with("_rate"), ~round(.x * 100, 1))
  )

# Render interactive table
render_reactable(
  summary_table,
  columns = list(
    complete_response_rate = colDef(
      name = "Complete Response (%)",
      format = colFormat(suffix = "%")
    ),
    partial_response_rate = colDef(
      name = "Partial Response (%)", 
      format = colFormat(suffix = "%")
    ),
    overall_response_rate = colDef(
      name = "Overall Response (%)",
      format = colFormat(suffix = "%")
    )
  ),
  defaultPageSize = 15
)
```

### SVG Plots with Interactivity

```{r interactive-svg}
# Create interactive scatter plot
interactive_plot <- iris %>%
  plot(
    mapping = list(
      x = "Sepal.Length",
      y = "Petal.Length", 
      color = "Species",
      tooltip = "paste('Species:', Species, '<br>Sepal Length:', Sepal.Length, '<br>Petal Length:', Petal.Length)"
    ),
    layers = list(
      list(type = "scatter", size = 3, alpha = 0.8)
    )
  )

# Convert to interactive HTML
as_html(interactive_plot)
```

## Inline and Report-Ready Visualizations

### Inline Plots for Text Integration

```{r inline-plots}
# Create small inline plots for reports
set.seed(123)
sample_data <- rnorm(200, mean = 50, sd = 10)

# Histogram inline plot
inline_hist <- add_inline_plot(
  sample_data,
  style = "Inline-Histogram",
  height = 0.4,
  color = "#3498DB"
)

inline_hist

# Violin inline plot
inline_violin <- add_inline_plot(
  sample_data,
  style = "Inline-Violin",
  height = 0.3,
  color = "#E74C3C"
)

inline_violin

# Point range inline plot
group_data <- tibble(
  group = LETTERS[1:5],
  mean = c(45, 52, 48, 55, 50),
  se = c(2.1, 1.8, 2.3, 1.9, 2.0)
)

inline_pointrange <- add_inline_pointrange(
  group_data,
  x = "group",
  y = "mean",
  ymin = "mean - se",
  ymax = "mean + se",
  height = 0.5
)

inline_pointrange
```

### Publication-Ready Styling

```{r publication-styling}
# Load publication-ready styles
# style_config <- load_style("Publication-Ready")

# Create publication-quality plot
publication_plot <- surv_data %>%
  filter(treatment %in% c("Control", "Treatment A")) %>%
  plot_km(
    time = "time",
    event = "event",
    treatment = "treatment"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 11, face = "bold")
  ) +
  labs(
    title = "Overall Survival: Control vs Treatment A",
    x = "Time (months)",
    y = "Survival Probability",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#E74C3C", "Treatment A" = "#3498DB"))

publication_plot
```

## Performance Optimization

### Caching for Large Datasets

```{r performance-caching}
# Enable caching for better performance
options(nightowl.cache = TRUE)

# Create large dataset
large_dataset <- tibble(
  group = sample(LETTERS[1:10], 50000, replace = TRUE),
  value = rnorm(50000),
  category = sample(c("X", "Y", "Z"), 50000, replace = TRUE)
)

# Time plot creation (first run - cache miss)
system.time({
  large_plot <- large_dataset %>%
    sample_n(1000) %>%  # Sample for demonstration
    plot(
      mapping = list(x = "group", y = "value", fill = "category"),
      layers = list(list(type = "boxplot"))
    )
})

# Second access (cache hit - faster)
system.time({
  large_plot_cached <- large_dataset %>%
    sample_n(1000) %>%
    plot(
      mapping = list(x = "group", y = "value", fill = "category"),
      layers = list(list(type = "boxplot"))
    )
})
```

## Advanced Statistical Visualizations

### Time Series and Longitudinal Data

```{r longitudinal-analysis}
# Create longitudinal dataset
longitudinal_data <- tibble(
  patient_id = rep(1:50, each = 6),
  visit = rep(c("Baseline", "Month 1", "Month 3", "Month 6", "Month 9", "Month 12"), 50),
  treatment = rep(sample(c("A", "B"), 50, replace = TRUE), each = 6),
  score = rnorm(300, mean = 75, sd = 15) + 
          rep(rnorm(50, 0, 5), each = 6) +  # Random patient effect
          c(0, -2, -5, -8, -10, -12)[rep(1:6, 50)]  # Treatment effect over time
) %>%
  mutate(
    visit = factor(visit, levels = c("Baseline", "Month 1", "Month 3", "Month 6", "Month 9", "Month 12"))
  )

# Longitudinal analysis plot
longitudinal_data %>%
  plot(
    mapping = list(
      x = "visit",
      y = "score",
      color = "treatment",
      group = "treatment"
    ),
    layers = list(
      list(type = "generic", geom = "ggplot2::stat_summary",
           fun = "mean", geom = "point", size = 3),
      list(type = "generic", geom = "ggplot2::stat_summary",
           fun = "mean", geom = "line", size = 1.2),
      list(type = "generic", geom = "ggplot2::stat_summary",
           fun.data = "mean_se", geom = "errorbar", width = 0.2)
    )
  ) +
  labs(
    title = "Treatment Response Over Time",
    subtitle = "Mean scores with standard errors",
    x = "Visit",
    y = "Clinical Score",
    color = "Treatment"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Dose-Response Relationships

```{r dose-response}
# Dose-response analysis
dose_data <- tibble(
  dose = rep(c(0, 10, 25, 50, 100, 200), each = 20),
  response = sapply(dose, function(d) {
    # Sigmoid dose-response curve with noise
    max_response <- 100
    ec50 <- 50
    hill_slope <- 1.5
    baseline <- 10
    
    expected_response <- baseline + (max_response - baseline) / (1 + (ec50/d)^hill_slope)
    rnorm(20, expected_response, 8)
  }) %>% as.numeric()
)

dose_data %>%
  plot(
    mapping = list(x = "dose", y = "response"),
    layers = list(
      list(type = "scatter", alpha = 0.6, size = 2),
      list(type = "smooth", method = "nls", 
           formula = "y ~ SSlogis(x, Asym, xmid, scal)",
           se = TRUE, color = "red")
    )
  ) +
  labs(
    title = "Dose-Response Relationship",
    subtitle = "Sigmoid curve fit with confidence bands",
    x = "Dose (mg)",
    y = "Response (%)"
  ) +
  theme_minimal()
```

## Customization and Styling

### Custom Color Palettes

```{r custom-colors}
# Define custom color palette
custom_colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b")

# Apply to visualization
iris %>%
  plot(
    mapping = list(
      x = "Sepal.Width",
      y = "Petal.Width",
      color = "Species"
    ),
    layers = list(list(type = "scatter", size = 3))
  ) +
  scale_color_manual(values = custom_colors[1:3]) +
  labs(title = "Custom Color Palette Example") +
  theme_minimal()
```

### Theme Customization

```{r theme-customization}
# Create custom theme
custom_theme <- theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", color = "#2C3E50"),
    plot.subtitle = element_text(size = 12, color = "#7F8C8D"),
    axis.title = element_text(size = 11, face = "bold"),
    legend.title = element_text(size = 10, face = "bold"),
    panel.grid.major = element_line(color = "#ECF0F1", size = 0.5),
    panel.grid.minor = element_blank()
  )

# Apply custom theme
mtcars %>%
  plot(
    mapping = list(x = "wt", y = "mpg", color = "factor(cyl)"),
    layers = list(
      list(type = "scatter", size = 3),
      list(type = "smooth", method = "lm", se = FALSE)
    )
  ) +
  custom_theme +
  labs(
    title = "Custom Theme Example",
    subtitle = "Weight vs Fuel Efficiency",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon",
    color = "Cylinders"
  )
```

## Best Practices

### 1. Consistent Styling
Always use consistent color schemes and themes across related visualizations.

### 2. Statistical Rigor
Include appropriate statistical tests and confidence intervals.

### 3. Clear Labeling
Provide descriptive titles, subtitles, and axis labels.

### 4. Performance Optimization
Enable caching for large datasets and repeated operations.

### 5. Interactive Elements
Leverage nightowl's interactive capabilities for exploratory analysis.

## Next Steps

Explore these advanced topics:

- **[Advanced Features](advanced-features.html)**: R6 classes and customization
- **[Performance Optimization](performance-optimization.html)**: Large dataset handling
- **[Real-World Examples](real-world-examples.html)**: Applied case studies

## Summary

nightowl provides a comprehensive statistical visualization toolkit that combines:

- **Clinical Focus**: Specialized functions for medical research
- **Statistical Integration**: Built-in statistical testing and modeling
- **Interactive Capabilities**: Modern web-based visualizations
- **Performance Optimization**: Efficient handling of large datasets
- **Publication Quality**: Professional styling and export options

The declarative approach and R6 architecture ensure reproducible, maintainable visualizations suitable for both exploratory analysis and publication-ready reports.